---
title: "Econ 491 Problem Set 1"
author: "Ramsey EL Lethy"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
---


## Question 1 (concept)[30p]

Explain whether each scenario is a classification or regression problem, and indicate
whether we are most interested in inference or prediction. Finally, provide n and p.

(a) We collect a set of data on the top 500 firms in the US. For each firm we record
profit, number of employees, industry and the CEO salary. We are interested in
understanding which factors affect CEO salary.

This is a regression problem, CEO Salary is quantitative

Predictors: Profit, # of Employees, Industry 

N : 500 Firms sampled 

We're more interested in inference, we're trying to understand which factors
influence CEO salary

 (b) We are considering launching a new product and wish to know whether it will be
a success or a failure. We collect data on 20 similar products that were previously
launched. For each product we have recorded whether it was a success or failure,
price charged for the product, marketing budget, competition price, and ten other
variables.

This is a Classfication problem, The target variable, product launch success is 
qualitative 

Predictors: Price charged for the product, marketing budget, comp price, ten other 
variables 

N : 20 Similar products 

We're trying to predict the success or failure of a new product given information about 
similar products 


(c) We are interested in predicting the % change in the USD/Euro exchange rate in
relation to the weekly changes in the world stock markets. Hence we collect weekly
data for all of 2012. For each week we record the % change in the USD/Euro, the
% change in the US market, the % change in the British market, and the % change
in the German market.

This is a regression problem, % Change is quantitative 

Predictors: % Change in US market, % Change in british market, 
% Change in German Market

N: 52 samples (1 years worth of weekly data)

We're trying to predict the percent change for a future week based on pased
information about the market, so this is a prediciton problem. 




```{r Part A & B}
college <- read.csv("College.csv", row.names = 1, stringsAsFactors = 1)
View(college)
```


```{r Part C.i}
summary(college)
```

```{r Part C.ii}
pairs(college[,1:10])
```



```{r Part C.iii}
attach(college)
plot(Private, Outstate, col = "purple")
```
```{r Part C.iv}
Elite <- rep("No", nrow(college))
Elite[Top10perc > 50] <- "Yes"
Elite <- as.factor(Elite)
college <- data.frame(college, Elite)
summary(college)

```


There are 78 Elite universities in this Dataset 


```{r Part C.iv cont.}
plot(Elite, Outstate)
```

```{r Part C.v}
par(mfrow = c(2, 2))
hist(Top10perc, breaks = 30)
hist(Books)
hist(Outstate, breaks = 100)
```